### 2022-10-01 chapters/benchmarking/experiment.tex
proofread : experiment.tex

We used the platform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708}, which is a large, flexible testbed for experiment-driven research that is spread out across all of France, to meet these needs. Grid5000 has several clusters that are made up of 4 to 124 identical machines with different configurations for each cluster. For our experiment, we looked at four groups. Our main criterion was the CPU configuration.
Table~\ref{table:g5k}  below, presents a description of the four clusters that have been chosen for our experiments.

@@

As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or \textsf{socket} to designate one of the two sockets and \textsf{PU} for the \emph{processing unit}.
For our study, we consider hyper-threads as distinct \textbf{PUs}.
Figure~\ref{fig:topo} shows the detailed topology of a node in the cluster \textsf{Dahu} as an example.

@@

To meet these criteria, we looked at the "state of the art" and found the most common benchmarks used to test the performance of hardware. We then chose the ones that could be scaled up.
Our candidate is \texttt{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925}: one of the most used benchmarks for \emph{HPC}. We used the applications (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
Furthermore we have used other applications to validate our results using more applications such as \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}.\footnote{\url{https://linux.die.net/man/1/sha256sum}}

\subsubsection{Metrics \& Measurement Tools}
Our metric for the accuracy of the test is the Standard deviation aka \textbf{STD} of the energy consumption. Therefore whether the tests consume more or less energy is out of our scope.
To study this variation we need first a tool to measure the energy consumption. For this we used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, which is a power monitoring toolkit that is based on \emph{Intel Running Average Power Limit} (RAPL) \cite{Khan:2018:RAE:3199681.3177754}. The advantage of PowerAPI is that it reports the Energy consumption of CPU and DRAM at a socket level.

Our testbeds are run with a minimal version of Debian \,9 (4.9.0 kernel version)\footnote{\url{https://github.com/grid5000/environments-recipes/blob/master/debian9-x64-min.yaml}} where we install Docker (version 18.09.5), which will be used to run the RAPL sensor and the benchmark itself.
The energy sensor collects RAPL reports and stores them in a remote \textsc{MongoDB} instance, allowing us to perform \emph{post-mortem} analysis in a dedicated environment.
Using Docker makes the deployment process easier on the one hand, and provides us with a built-in control group encapsulation of the conducted tests on the other hand.
This allows \textsc{PowerAPI} to measure all the running containers, even the RAPL sensor consumption, as it is isolated in a container.

Every experiment is conducted on $100$ iterations, on multiple nodes and using the 3 NPB benchmarks we mentioned, with a warmup phase of 10 iterations for each experiment.
In most cases, we were seeking to evaluate the \emph{STandard Deviation} (STD), which is the most representative factor of the energy variation.
We tried to be very careful, while running our experiments, not to fall into the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
As we study the STD difference of measurements we observed from empirical experiments, we use the bootstrap method~\cite{efron2000bootstrap} to randomly build multiple subsets of data from the original dataset, and we draw the STD density of those sets, as illustrated in Figure~\ref{fig:docker}.
Given the space constraints, we report on aggregated results for nodes, benchmarks and workloads.
However, the raw data we collected is available through the public repository we published.\footnote{\url{https://github.com/anonymous-data/Energy-Variation}}
We believe this can help to achieve better and more reliable comparisons.
We mainly consider 3 different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium and high CPU usage when analyzing the effect of the studied parameters, respectively.
These workloads reflect the ratio of used PU count to the total available PU.

@@

To answer this question we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running the same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
While Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.

@@

\subsubsection{\textsc{RQ}~1: Benchmarking Protocol}
To achieve a robust and reproducible experiment, practitioners often tend to repeat their tests multiple times, in order to analyze the related performance indicators, such as execution time, memory consumption or energy consumption.
We, therefore, aim to study the benchmarking protocol to identify how to efficiently iterate the tests to capture a trustable energy consumption evaluation.
% This process of repeating executions should not be random in order to avoid different results because of the CPU behavior, the cache data, the memory size and plenty of other aspects.

In this first experiment, we investigate if changing the testing protocol affects the energy variation.
To achieve this, we considered 3 execution modes:
In the \textsf{"normal"} mode, we iteratively run the benchmark 100 times without any extra command, while the \textsf{"sleep"} mode suspends the execution script for 60\,seconds between iterations.
Finally, the \textsf{"reboot"} mode automatically reboots the machine after each iteration.
The difference between the \textsf{normal} and \textsf{sleep} modes intends to highlight that the CPU needs some rest before starting another iteration, especially for an intense workload.
Putting the CPU into sleep for several seconds could give it some time to reach a lower frequency state or/and reduce its temperature, which could have an impact on the energy variation.
The \textsf{reboot} mode, on the other hand, is the most straightforward way to reset the machine state after every iteration.
It could also be beneficial to reset the CPU frequency and temperature, the stored data, the cache or the CPU registries.
However, the reboot task takes a considerable amount of time, so rebooting the node after every single operation is not the fastest nor the most eco-friendly solution, but it deserves to be checked to investigate if it effectively enhances the overall energy variation or not.

Figure~\ref{fig:running-process} reports on $300$ aggregated executions of the  benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}, on 4 machines of the cluster \textsf{Dahu} (cf. Table~\ref{table:g5k}) for different workloads.
We note that the results have been executed with different datasets sizes (\textsf{B}, \textsf{C} and \textsf{D} for a single process, 50\,\% and 100\,\% respectively) to remedy to the brief execution times at high workloads for small datasets.
This justifies the scale differences in reported energy consumption between the 3 modes in Figure~\ref{fig:running-process}.
As one can observe, picking one of these strategies does not have a strong impact on the energy variation for most workloads.
In fact, all the strategies seem to exhibit the same variation with all the workloads we considered---\emph{i.e.}, the STD is tightly close between the three modes.
The only exception is the \textsf{reboot} mode at 100\,\% load, where the STD is 150\,\% times worst, due to an important amount of outliers.
This goes against our expectations, even when setting a warm-up time after reboot to stabilize the OS.

@@

In Figure~\ref{fig:reboot}, we study the standard deviation of the three modes by constituting $5,000$ random 30-iterations sets from the previous executions set and we compute the STD in each case, considering mainly the 100\,\% workload as the STD was 150\,\% higher for the \textsf{reboot} mode with that load.
We can observe that the considerable amount of outliers in the \textsf{reboot} mode is not negligible, as the STD density is higher than the two other modes.
This makes the \textsf{reboot} mode less appropriate for the energy variation at high workloads.

@@

    To answer \textsc{RQ~1}, we conclude that the benchmarking protocol \textbf{partially affects} the energy variation, as highlighted by the \textsf{reboot} mode results in high workloads.

@@

\subsubsection{\textsc{RQ}~2: Processor Features}
The C-states provide the ability to switch the CPU between more or less consuming states upon activities.
Turning the C-states on or off have been subject of many discussions~\cite{5463056}, because of its dynamic frequency mechanism but, to the best of our knowledge, there have been no fully conducted C-states behavior analysis on CPU energy variation.

We intend to investigate how much the energy consumption varies when disabling the C-states (thus, keeping the CPU in the \texttt{C0} state) and at which workload.
Figure~\ref{fig:c-states} depicts the results of the experiments we executed on three nodes of the cluster \textsf{Dahu}.
On each node, we ran the same set of benchmarks with two modes: \textsf{C-states on}, which is the default mode, and \textsf{C-states off}.
Each iteration includes 100 executions of the same benchmark at a given workload, with three workload levels.
We note that our results have been confirmed with the benchmarks \texttt{LU}, \texttt{CG} and \texttt{EP}.

We can see the effect that has the \textsf{C-states off} mode when running a single-process application/benchmark.
The energy consumption varies 5~times less than the default mode.
In this case, only one CPU core is used among $2\times16$ physical cores.
The other cores are switched to a low consumption state when C-states are on, the switching operation causes an important energy consumption difference between the cores, and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
On the other hand, switching off the C-states would keep all the cores---even the unused ones---at a high-frequency usage.
This highly reduces the variation, but causes up to 50\,\% of extra energy consumption in this test ($Mean_{C-states-off}=11,665 mJ$,$Mean_{C-states-on}=7,641 mJ$).

At a 100\,\% workload, disabling the C-states seems to not affect the total energy consumption nor its variation.
All the cores are used at 100\,\% and the C-states module would have no effect, as the cores are not idle.
The same reason would apply for the 50\,\% load, as the hyper-threading is active on all cores, thus causing the usage of most of them.
For single process workloads, disabling the C-states causes the process to consume 50\,\% more energy as reported in Figure~\ref{fig:c-states}, but reduces the variation by 5~times compared to the \textsf{C-states on} mode.
This leads to mainly two questions: Can a process pinning method reduce/increase the energy variation? And, how does the energy consumption variation evolve at different PU usage levels?

@@

\paragraph{Cores Pinning}
To answer the first question, we repeated the previous test at 50\,\% workload.
In this experiment, we considered three cores usage strategies, the first one (\textsf{S1}) would pin the processes on all the PU of one of the two sockets (including hyper-threads), so it will be used at 100\,\%, and leave the other CPU idle.
The second strategy (\textsf{S2}) splits the workload on the two sockets so each CPU will handle 50\,\% of the load.
In this strategy, we only use the core PU and not the hyper-threads PU, so every process would not share its core usage (all the cores are being used).
The third strategy (\textsf{S3}) consists also of splitting the workload between the two sockets, but considering the usage of the hyper-threads on each core---\emph{i.e.}, half of the cores are being used over the two CPU.
Figure~\ref{fig:cores-pinning} reports on the energy consumption of the three strategies when running the benchmark \texttt{CG} on the cluster \textsf{Dahu}.
We can notice the big difference between these three execution modes that we obtained only by changing the PU pinning method (that we acknowledged with more than 100 additional runs over more than 30 machines and with the benchmarks \texttt{LU} and \texttt{EP}).
For example, \textsf{S2} is the least power-consuming strategy.
We argue that the reason is related to the isolation of every process on a single physical core, reducing the context switch operations.
In the first and third strategies, $32$ processes are being scheduled on $16$ physical cores using the hyper-threads PU, which will introduce more context switching, and thus more energy consumption.

@@

We note that even if the first and third strategies are very similar (both use hyper-threads, but only on one CPU for the first and on two CPU for the third), the gap between them is considerable variation-wise, as the variation is 30 times lower in the first strategy ($STD_{S1}=116 mJ$,$STD_{S3}=3,452 mJ$).
This shows that the usage of the hyper-threads technology is not the main reason behind the variation, the first strategy has even less variation than the second one and still uses the hyper-threading.

The reason for the \textsf{S1} low energy consumption is that one of the two sockets is idle and will likely be in a lower power P-state, even with the disabled C-states.
The \textsf{S2} case reports on low energy consumption because by distributing the threads across all the cores, it completes the task faster than in the other cases.
Hence, it consumes less energy.
The \textsf{S3} is a highly consuming strategy because both sockets are being used, but only half the cores are active.
This means that we pay the energy cost for both sockets being operational and for the experiments taking longer to run because of the recurrent context switching.

Our hypothesis regarding the worst results that we observed when using the third strategy is the recurrent context switching, added to the OS scheduling that could reschedule processes from a socket to another, which invalids the cache usage as a process can not take profit of the socket local L3 cache when it moves from a CPU to another (cf. Figure~\ref{fig:topo}).

@@

Table~\ref{table:corespinning} reports on additional aggregated results for the STD comparison on four other nodes of the cluster \textsf{Dahu} at 50\,\%, with the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
The CPU usage strategy \textsf{S1} is by far the experimentation mode that gave the least variation.
The STD is almost 5 times better than the strategy \textsf{S2}, but is up to 10\,\% more energy consuming ($Mean_{S1}=4469 mJ$, $Mean_{S2}=4016 mJ$).
On the other hand, the strategy \textsf{S3} is the worst, where the energy consumption can be up to 5 times higher than the strategy \textsf{S2} ($Mean_{S2}=4016 mJ$, $Mean_{S3}=21645 mJ$) and the variation is much worst (30 times compared to the first strategy).
These results allow us to have a better understanding of the different processes-to-PU pinning strategies, where isolating the workload on a single CPU is the best strategy.
Using the hyper-threads PU on multiple sockets seems to be a bad recommendation, while keeping the hyper-threading enabled on the machine is not problematic, as long as the processes are correctly pinned on the PU.
Our experiments show that running one hyper-thread per core is not always the best to do, at the opposite of the claims of~\cite{marathe_empirical_2017_m}.

@@

\paragraph{Processes Threshold}
To answer the second question regarding the evolution of the energy variation at different levels of CPU usage, we varied the used PU's count to track the EV evolution.
Figure~\ref{fig:process-tresh} compares the aggregated energy variation when the C-states are on and off using 2, 4 and 8 processes for the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
This figure confirms that disabling the CPU C-states does not decrease the variation for all the workloads, as we can clearly observe, the variation is increasing along with the number of processes.
When running only 2 processes, turning off the C-states reduces the STD up to 6~times, but consumes 20\,\% more energy ($Mean_{C-states-on}=10,334 mJ$, $Mean_{C-states-off}=12,594 mJ$).
This variation is 4~times lower when running 4 processes and almost equal to the \textsf{C-states on} mode when running 8 processes.
In fact, running more processes implies using more CPU cores, which reduces the idle cores count, so the cores will more likely stay at a higher consumption state even if the C-states mechanism is on.

In our case, using 4 PU reduces the variation by 4 times and consumes almost the same energy as keeping the C-states mechanism on ($Mean_{C-states-on}=7,048 mJ$, $Mean_{C-states-off}=7,119 mJ$).
This case would be the closest to reality as we do not want to increase the energy consumption while reducing the variation, but using a lower number of PU still results in less variation, even if it increases the overall energy consumption.

@@

We note that disabling the C-states is not recommended in production environments, as it introduces extra energy consumption for low workloads (around 50\,\% in our case for a single process job).
However, our goal is not to optimize the energy consumption, but to minimize the energy variation.
Thus, disabling the C-states is very important to stabilize the measurements in some cases when the variation matters the most.
Comparing the energy consumption of two algorithms or two versions of a software system is an example of a use case benefiting from this recommendation.

\paragraph{Turbo Boost}
The Turbo Boost---also known as \emph{Dynamic Overclocking}---is a feature that has been incorporated in Intel CPU since the Sandy~Bridge micro-architecture, and is now widely available on all of the Core\,i5, Core\,i7, Core\,i9 and Xeon series.
It automatically raises some of the CPU cores operating frequency for short periods, and thus boosts performances under specific constraints.
When demanding tasks are running, the operating system decides on using the highest performance state of the processor.

Disabling or enabling the Turbo~Boost has a direct impact on the CPU frequency behavior, as enabling it allows the CPU to reach higher frequencies to execute some tasks for a short period.
However, its usage does not have a trivial impact on the energy variation.
Acun~\emph{et~al.}~\cite{acun_variation_2016} tried to track the Turbo~Boost impact on the Ivy~Bridge and the Sandy~Bridge architectures.
They concluded that it is one of the main responsible for the energy variation, as it increases the variation from 1\,\% to 16\,\%.
In our study, we included a Turbo~Boost experiment in our testbed, to check this property on the recent Xeon Gold processors, covering various workloads.

The experiment we conducted showed that disabling the Turbo Boost does not exhibit any considerable positive or negative effect on the energy variation.
Table~\ref{table:turboboost} compares the STD when enabling/disabling the Turbo~Boost, where the columns are a combination of workload and benchmark.
We only got some minor measurement differences when switching on and off the Turbo~Boost, and were in favor or against the usage of the Turbo~Boost while repeating tests, considering multiple nodes and benchmarks.
This behavior is mainly related to the \emph{thermal design power} (TDP), especially at high workloads executions.
When a CPU is used at its maximum capacity, the cores would be heating up very fast and would hit the maximum TDP limit.
In this case, the Turbo~Boost cannot offer more power to the CPU because of the CPU thermal restrictions.
At lower workloads, the tests we conducted proved that the Turbo~Boost is not one of the main reasons for the energy variation.
The variation difference is barely noticeable when disabling the Turbo~Boost, which cannot be considered as a result of the OS activity and the measurement error margin.
We cannot affirm that the Turbo~Boost does not have an impact on all the CPUs, as we only tested on two recent Xeon CPUs (clusters \textsf{Chetemi} and \textsf{Dahu}).
We confirmed our experiments on these machines 100 times at 5\,\%, 25\,\%, 50\,\% and 100\,\% workloads.

>>>>

We used the platform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708}, a large, flexible testbed for experiment-driven research spread across France, to meet these needs. Grid5000 has several clusters comprising 4 to 124 identical machines with different configurations for each cluster. We looked at four groups for our experiment, and our main criterion was the CPU configuration.
Table~\ref{table:g5k} below presents a description of the four clusters that have been chosen for our experiments.

@@

As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or \textsf{socket} to designate one of the two sockets and \textsf{PU} for the \emph{processing unit}.
Our study considers hyper-threads as distinct \textbf{PUs}.
Figure~\ref{fig:topo} shows the detailed topology of a node in the cluster \textsf{Dahu} as an example.

@@

To meet these criteria, we looked at the "state of the art" and found the most common benchmarks used to test hardware performance. We then chose the ones that could be scaled up.
Our candidate is \texttt{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925}: one of the most used benchmarks for \emph{HPC}. We used the applications (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
Furthermore we have used other applications to validate our results using more applications such as \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}.\footnote{\url{https://linux.die.net/man/1/sha256sum}}

\subsubsection{Metrics \& Measurement Tools}
Our metric for the accuracy of the test is the Standard deviation aka \textbf{STD} of the energy consumption. Therefore whether the tests consume more or less energy is out of our scope.
We first need a tool to measure energy consumption to study this variation. For this we used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, which is a power monitoring toolkit that is based on \emph{Intel Running Average Power Limit} (RAPL) \cite{Khan:2018:RAE:3199681.3177754}. The advantage of PowerAPI is that it reports the Energy consumption of CPU and DRAM at a socket level.

Our testbeds are run with a minimal version of Debian \,9 (4.9.0 kernel version)\footnote{\url{https://github.com/grid5000/environments-recipes/blob/master/debian9-x64-min.yaml}} where we install Docker (version 18.09.5), which will be used to run the RAPL sensor and the benchmark itself.
The energy sensor collects RAPL reports and stores them in a remote \textsc{MongoDB} instance, allowing us to perform \emph{post-mortem} analysis in a dedicated environment.
Using Docker makes the deployment process easier on the one hand and provides us with a built-in control group encapsulation of the conducted tests on the other hand.
Docker allows \textsc{PowerAPI} to measure all the running containers, even the RAPL sensor consumption, as it is isolated in a container.

Every experiment is conducted on $100$ iterations, on multiple nodes and using the 3 NPB benchmarks we mentioned, with a warmup phase of 10 iterations for each experiment.
In most cases, we sought to evaluate the \emph{Standard Deviation} (STD), which is the most representative factor of the energy variation.
While running our experiments, we tried not to fall into the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
As we study the STD difference of measurements we observed from empirical experiments, we use the bootstrap method~\cite{efron2000bootstrap} to randomly build multiple subsets of data from the original dataset, and we draw the STD density of those sets, as illustrated in Figure~\ref{fig:docker}.
Given the space constraints, we report on aggregated results for nodes, benchmarks, and workloads.
However, the raw data we collected is available through the public repository we published.\footnote{\url{https://github.com/anonymous-data/Energy-Variation}}
We believe this can help to achieve better and more reliable comparisons.
We mainly consider three different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium, and high CPU usage when analyzing the effect of the studied parameters, respectively.
These workloads reflect the ratio of used PU count to the total available PU.

@@

To answer this question, we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on three nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running the same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
While Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.

@@

\subsubsection{\textsc{RQ}~1: Benchmarking Protocol}
To achieve a robust and reproducible experiment, practitioners often repeat their tests multiple times to analyze the related performance indicators, such as execution time, memory consumption, or energy consumption.
We, therefore, aim to study the benchmarking protocol to identify how to efficiently iterate the tests to capture a trustable energy consumption evaluation.
% This process of repeating executions should not be random in order to avoid different results because of the CPU behavior, the cache data, the memory size and plenty of other aspects.

In this first experiment, we investigate if changing the testing protocol affects the energy variation.
To achieve this, we considered three execution modes:
In the \textsf{"normal"} mode, we iteratively run the benchmark 100 times without any extra command, while the \textsf{"sleep"} mode suspends the execution script for 60\,seconds between iterations.
Finally, after each iteration, the \textsf{"reboot"} mode automatically reboots the machine.
The difference between the \textsf{normal} and \textsf{sleep} modes highlights that the CPU needs rest before starting another iteration, especially for an intense workload.
Putting the CPU into sleep for several seconds could give it time to reach a lower frequency state or/and reduce its temperature, which could impact the energy variation.
On the other hand, the \textsf{reboot} mode is the most straightforward way to reset the machine state after every iteration.
It could also be beneficial to reset the CPU frequency and temperature, the stored data, the cache, or the CPU registries.
However, the reboot task takes a considerable amount of time, so rebooting the node after every single operation is not the fastest nor the most eco-friendly solution. However, it deserves to be checked to investigate if it effectively enhances the overall energy variation or not.

Figure~\ref{fig:running-process} reports on $300$ aggregated executions of the  benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}, on 4 machines of the cluster \textsf{Dahu} (cf. Table~\ref{table:g5k}) for different workloads.
We note that the results have been executed with different datasets sizes (\textsf{B}, \textsf{C} and \textsf{D} for a single process, 50\,\% and 100\,\% respectively) to remedy the brief execution times at high workloads for small datasets.
This justifies the scale differences in reported energy consumption between the three modes in Figure~\ref{fig:running-process}.
One can observe that picking one of these strategies does not strongly impact the energy variation for most workloads.
All the strategies exhibit the same variation with all the workloads we considered---\emph{i.e.}, the STD is tightly close between the three modes.
The only exception is the \textsf{reboot} mode at 100\,\% load, where the STD is 150\,\% times worst due to a substantial amount of outliers.
This goes against our expectations, even when setting a warm-up time after reboot to stabilize the OS.

@@

In Figure~\ref{fig:reboot}, we study the standard deviation of the three modes by constituting $5,000$ random 30-iterations sets from the previous executions set, and we compute the STD in each case, considering mainly the 100\,\% workload as the STD was 150\,\% higher for the \textsf{reboot} mode with that load.
We can observe that the considerable amount of outliers in the \textsf{reboot} mode is not negligible, as the STD density is higher than the two other modes.
This makes the \textsf{reboot} mode less appropriate for the energy variation at high workloads.

@@

    To answer \textsc{RQ~1}, we conclude that the benchmarking protocol \textbf{partially affects} the energy variation, as highlighted by the \textsf{reboot} mode, results in high workloads.

@@

\subsubsection{\textsc{RQ}~2: Processor Features}
The C-states allow switching the CPU between more or less consuming states upon activities.
Turning the C-states on or off have been the subject of many discussions~\cite{5463056} because of its dynamic frequency mechanism, but, to the best of our knowledge, there has been no entirely conducted C-states behavior analysis on CPU energy variation.

We intend to investigate how much the energy consumption varies when disabling the C-states (thus, keeping the CPU in the \texttt{C0} state) and at which workload.
Figure~\ref{fig:c-states} depicts the results of the experiments we executed on three nodes of the cluster \textsf{Dahu}.
On each node, we ran the same benchmarks with two modes: \textsf{C-states on}, which is the default mode, and \textsf{C-states off}.
Each iteration includes 100 executions of the same benchmark at a given workload, with three workload levels.
We note that our results have been confirmed with the benchmarks \texttt{LU}, \texttt{CG} and \texttt{EP}.

We can see the effect that has the \textsf{C-states off} mode when running a single-process application/benchmark.
The energy consumption varies 5~times less than the default mode.
In this case, only one CPU core is used among $2\times16$ physical cores.
The other cores are switched to a low consumption state when C-states are on. The switching operation causes a vital energy consumption difference between the cores and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
On the other hand, switching off the C-states would keep all the cores---even the unused ones---at a high-frequency usage.
This highly reduces the variation, but causes up to 50\,\% of extra energy consumption in this test ($Mean_{C-states-off}=11,665 mJ$,$Mean_{C-states-on}=7,641 mJ$).

At a 100\,\% workload, disabling the C-states seems not to affect the total energy consumption nor its variation.
All the cores are used at 100\,\%, and the C-states module would have no effect, as the cores are not idle.
The same reason would apply for the 50\,\% load, as the hyper-threading is active on all cores, thus causing the usage of most of them.
For single process workloads, disabling the C-states causes the process to consume 50\,\% more energy as reported in Figure~\ref{fig:c-states}, but reduces the variation by 5~times compared to the \textsf{C-states on} mode.
This leads to two questions: Can a process pinning method reduce/increase the energy variation? Moreover, how does the energy consumption variation evolve at different PU usage levels?

@@

\paragraph{Cores Pinning}
To answer the first question, we repeated the previous test at 50\,\% workload.
In this experiment, we considered three cores usage strategies, the first one (\textsf{S1}) would pin the processes on all the PU of one of the two sockets (including hyper-threads), so it will be used at 100\,\%, and leave the other CPU idle.
The second strategy (\textsf{S2}) splits the workload on the two sockets so each CPU will handle 50\,\% of the load.
In this strategy, we only use the core PU and not the hyper-threads PU, so every process would not share its core usage (all the cores are being used).
The third strategy (\textsf{S3}) also involves splitting the workload between the two sockets, but with the hyper-threads on each core used, i.e., half of the cores are used across the two CPUs.

Figure~\ref{fig:cores-pinning} reports on the energy consumption of the three strategies when running the benchmark \texttt{CG} on the cluster \textsf{Dahu}.
We can notice the big difference between these three execution modes that we obtained only by changing the PU pinning method (that we acknowledged with more than 100 additional runs over more than 30 machines and with the benchmarks \texttt{LU} and \texttt{EP}).
For example, \textsf{S2} is the least power-consuming strategy.
We argue that the reason is related to the isolation of every process on a single physical core, reducing the context switch operations.
In the first and third strategies, $32$ processes are being scheduled on $16$ physical cores using the hyper-threads PU, which will introduce more context switching, thus, more energy consumption.

@@

Even though the first and third strategies are very similar (both use hyper-threads, but only on one CPU for the first and two CPUs for the third), the variation gap between them is significant, as the variation in the first strategy is 30 times lower ($STD_{S1}=116 mJ$,$STD_{S3}=3,452 mJ$).
This shows that the hyper-threads technology is not the main reason behind the variation; the first strategy has even less variation than the second one and still uses the hyper-threading.

The reason for the \textsf{S1} low energy consumption is that one of the two sockets is idle and will likely be in a lower power P-state, even with the disabled C-states.
The \textsf{S2} case reports low energy consumption because distributing the threads across all the cores completes the task faster than in the other cases.
Hence, it consumes less energy.
The \textsf{S3} is a highly consuming strategy because both sockets are used, but only half the cores are active.
This means that we pay the energy cost for both sockets being operational and for the experiments taking longer to run because of the recurrent context switching.

Our hypothesis for the worst results observed when using the third strategy is that recurrent context switching, which is added to the OS scheduling and can reschedule processes from one socket to another, invalidates cache usage because a process cannot benefit from the socket local L3 cache when it moves from one CPU to another. (cf. Figure~\ref{fig:topo}).

@@

Table~\ref{table:corespinning} reports on additional aggregated results for the STD comparison on four other nodes of the cluster \textsf{Dahu} at 50\,\%, with the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
The CPU usage strategy \textsf{S1} is by far the experimentation mode that gave the least variation.
The STD is almost 5 times better than the strategy \textsf{S2}, but is up to 10\,\% more energy consuming ($Mean_{S1}=4469 mJ$, $Mean_{S2}=4016 mJ$).
On the other hand, the strategy \textsf{S3} is the worst, where the energy consumption can be up to 5 times higher than the strategy \textsf{S2} ($Mean_{S2}=4016 mJ$, $Mean_{S3}=21645 mJ$) and the variation is much worst (30 times compared to the first strategy).
These results allow us to understand better the different processes-to-PU pinning strategies, where isolating the workload on a single CPU is the best strategy.
Using the hyper-threads PU on multiple sockets seems to be a bad recommendation. Keeping the hyper-threading enabled on the machine is not problematic as long as the processes are correctly pinned on the PU.
Our experiments show that running one hyper-thread per core is not always the best to do, opposite the claims of~\cite{marathe_empirical_2017_m}.

@@

\paragraph{Processes Threshold}
To answer the second question regarding the evolution of the energy variation at different levels of CPU usage, we varied the used PU's count to track the EV evolution.
Figure~\ref{fig:process-tresh} compares the aggregated energy variation when the C-states are on and off using 2, 4 and 8 processes for the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
This figure confirms that disabling the CPU C-states does not decrease the variation for all the workloads; as we can clearly observe, the variation is increasing along with the number of processes.
When running only 2 processes, turning off the C-states reduces the STD up to 6~times, but consumes 20\,\% more energy ($Mean_{C-states-on}=10,334 mJ$, $Mean_{C-states-off}=12,594 mJ$).
This variation is 4~times lower when running four processes and almost equal to the \textsf{C-states on} mode when running eight processes.
In fact, running more processes implies using more CPU cores, which reduces the idle cores count, so the cores will more likely stay at a higher consumption state even if the C-states mechanism is on.

In our case, using 4 PU reduces the variation by four times and consumes almost the same energy as keeping the C-states mechanism on ($Mean_{C-states-on}=7,048 mJ$, $Mean_{C-states-off}=7,119 mJ$).
This case would be the closest to reality as we do not want to increase energy consumption while reducing the variation. However, using a lower number of PU still results in less variation, even if it increases the overall energy consumption.

@@

We note that disabling the C-states is not recommended in production environments, as it introduces extra energy consumption for low workloads (around 50\,\% in our case for a single process job).
However, our goal is not to optimize energy consumption but to minimize the energy variation.
Thus, disabling the C-states is very important to stabilize the measurements in some cases when the variation matters the most.
Comparing the energy consumption of two algorithms or two software system versions is an example of a use case benefiting from this recommendation.

\paragraph{Turbo Boost}
The Turbo Boost---also known as \emph{Dynamic Overclocking}---is a feature that has been incorporated in Intel CPU since the Sandy~Bridge micro-architecture, and is now widely available on all of the Core\,i5, Core\,i7, Core\,i9, and Xeon series.
It automatically raises some of the CPU cores operating frequency for short periods, thus boosting performance under specific constraints.
When demanding tasks run, the operating system decides to use the processor's highest performance state.

Disabling or enabling the Turbo~Boost  impacts the CPU frequency behavior directly, as enabling it allows the CPU to reach higher frequencies to execute some tasks for a short period.
However, its usage does not significantly impact the energy variation.
Acun~\emph{et~al.}~\cite{acun_variation_2016} tried to track the Turbo~Boost impact on the Ivy~Bridge and the Sandy~Bridge architectures.
They concluded that it is one of the main responsible for the energy variation, as it increases the variation from 1\,\% to 16\,\%.
In our study, we included a Turbo~Boost experiment in our testbed to check this property on the recent Xeon Gold processors, covering various workloads.

The experiment showed that disabling the Turbo Boost does not exhibit any considerable positive or negative effect on the energy variation.
Table~\ref{table:turboboost} compares the STD when enabling/disabling the Turbo~Boost, where the columns combine workload and benchmark.
We only had minor measurement differences when switching on and off the Turbo~Boost. We favored or against using the Turbo~Boost while repeating tests, considering multiple nodes and benchmarks.
This behavior is mainly related to the \emph{thermal design power} (TDP), especially at high workloads.
When a CPU is used at its maximum capacity, the cores would heat up very fast and hit the maximum TDP limit.
In this case, the Turbo~Boost cannot offer more power to the CPU because of the CPU thermal restrictions.
At lower workloads, the tests we conducted proved that the Turbo~Boost is not one of the main reasons for the energy variation.
The variation difference is barely noticeable when disabling the Turbo~Boost, which cannot be considered due to the OS activity and the measurement error margin.
We cannot affirm that the Turbo~Boost does not have an impact on all the CPUs, as we only tested on two recent Xeon CPUs (clusters \textsf{Chetemi} and \textsf{Dahu}).
We confirmed our experiments on these machines 100 times at 5\,\%, 25\,\%, 50\,\% and 100\,\% workloads.
